The Changing Landscape of Cystic Fibrosis Related Diabetes Risk Factors Between 2003-2018

### 2025-06-15 Christine Chan/CF Registry/WNAR/wnar_presentation.qmd
WNAR slides done

---
title: "The Changing Landscape of Cystic Fibrosis Related Diabetes Risk Factors Between 2003-2018"
author: "Tim Vigers"
date: "June 17, 2025"
date-format: long
format: 
  revealjs:
    scrollable: true
    embed-resources: true
editor: source
bibliography: wnar_references.bib
csl: american-medical-association.csl
---

@@

- Median predicted survival increased from 33 to 44 years between 2003-2018.[@cysticfibrosisfoundationCysticFibrosisFoundation2018]
- We expect to see even greater gains with the approval of new highly effective *CFTR* modulator therapies. [@Bessonova731; @middletonElexacaftorTezacaftorIvacaftor2019]
- However, despite these advances CFRD remains prevalent and there have been few large-scale studies examining changes in the epidemiology of CFRD over time.[@moranClinicalCareGuidelines2010; @marshallEpidemiologyCysticFibrosisrelated2005; @adlerHyperglycemiaDeathCystic2011; @chamnanDiabetesDeterminantMortality2010]

@@

## Known Risk Factors (cont.)

- Other studies have identified additional potential risk factors, but these are the most well-established.
- These variables were also readily available in the United States Cystic Fibrosis Foundation Patient Registry (CFFPR) and have been collected consistently since 2003.
- Very few model selection issues (yay!)

@@

## Participant Demographics in 2018 {.scrolling}

```{r}
#| message: false
load("./Data_Clean/survival_dataset.RData")
annual %>%
  filter(ReviewYear == 2018) %>%
  select(
    CFRD, Age_YrEnd, Gender, Race, Hispanicrace, bmi_perc,
    bmi_cat, MutClass, tube_feeding, liver_disease,
    A_FEV1_pct_predicted, A_FVC_pct_predicted, NumPulmExacerbation,
    A_corticosteroids1, A_IsOnEnzymes, A_Vx770, A_VX809comb
  ) %>%
  tbl_summary(
    by = CFRD, missing_text = "Missing",
    type = list(A_corticosteroids1 ~ "categorical"),
    statistic = list(c(
      "Age_YrEnd", "bmi_perc", "A_FEV1_pct_predicted",
      "A_FVC_pct_predicted"
    ) ~ "{mean} ({sd})")
  ) %>%
  add_p(test = list(c(
    "Age_YrEnd", "bmi_perc", "A_FEV1_pct_predicted",
    "A_FVC_pct_predicted"
  ) ~ "t.test")) %>%
  add_overall() %>%
  separate_p_footnotes()
```

@@

```{r}
#| fig-width: 16
#| fig-height: 9
# KM plot
fit <- survfit(
  Surv(time = tstop, event = as.numeric(CFRD) == 2) ~ age_group,
  data = annual %>%
    group_by(eDWID) %>%
    slice_max(ReviewYear)
)
# Visualize with survminer
ggsurvplot(fit,
  data = annual, risk.table = T, cumevents = T, palette = "simpsons",
  conf.int = T
)
```

@@

## Data Setup Notes {.smaller}

- Because we don't have an exact date of CFRD diagnosis, the only information we have is that the diagnosis occurred between January 1st of one year and January 1st of the next.
  - In other words, data are interval censored.
- Intervals are on a discrete grid (i.e. start and stop times are the same for all participants in a given year).
- Data include time-varying covariates such as age group, BMI category, etc.
- We categorized variables such as BMI and FEV1 mostly for the sake of interpretability for a clinical audience.
  - For BMI in particular though, we were not sure that the effect would be linear and categorization allowed for more flexible models.

@@

## Two good options

1. Joint modeling of longitudinal and survival data.[@hendersonJointModellingLongitudinal2000]
  - The gold standard, do this if you can.
  - Complex, requires somewhat specialized knowledge.
  - Ask Elizabeth for help with this.

2. Time-dependent Cox models
  - More appropriate for external covariates (measures that vary independent of the failure time).
  - Or baseline covariates where the effect changes over time.[@zhangTimevaryingCovariatesCoefficients2018]

## Time Dependent Covariates and Coefficients

@@

## Time Dependent Covariates and Coefficients

@@

```{r}
data(cancer)
vet2 <- survSplit(Surv(time, status) ~ .,
  data = cancer, cut = c(90, 180), episode = "tgroup", id = "id"
)
vfit2 <- coxph(Surv(tstart, time, status) ~ age + sex + ph.ecog +
  pat.karno:strata(tgroup), data = vet2)
vfit2
```

@@

```{r}
annual_split <- survSplit(Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2) ~ .,
  data = annual, cut = c(2192, 4018), episode = "tgroup"
)
cox_res2 <- coxph(
  Surv(time = tstart, time2 = tstop, event) ~
    age_group:strata(tgroup) + Gender + Race + Hispanicrace + bmi_cat + 
    MutClass:strata(tgroup) + A_IsOnEnzymes:strata(tgroup) + 
    tube_feeding:strata(tgroup) + liver_disease:strata(tgroup) + 
    ppFEV1_cat:strata(tgroup) + NumPulmExacerbation:strata(tgroup) + 
    A_corticosteroids1:strata(tgroup),
  data = annual_split, cluster = eDWID
)
```

@@

@@

@@

- Is this even worth including? From a statistics point of view it's not that interesting.

@@

- The middle age group is no longer at increased risk compared to the youngest.
- Female and Black people with CF remain at increased risk of developing CFRD.
- People with CF who are in the "undernourished" BMI category are no longer at increased risk, but those who are overweight are.
- Other risk factors appear relatively stable over time.
- Ivcaftor may reduce the risk of CFRD, although combination therapy does not appear to have an effect.

@@

- Methods available for interval censored data in the `survival` R package[@therneauPackageSurvivalAnalysis2024] are slightly limited.
  - Other packages such as `icenReg`[@anderson-bergmanIcenRegRegressionModels2017] may be able to fit more flexible models.
- Categorizing variables is not always ideal, and there is some concern about loss of information.
  - Currently we are working on extending this analysis using joint modeling for more flexibility.
- Data end in 2018 but Trikafta (a triple combination therapy of Elexacaftor, Ivacaftor, and Tezacaftor) was approved in 2019.
  - Need to study later generations of CFTR modulators.
- Registry data has natural limitations because each center must track and report their own data.

## Special Thanks To

- The Cystic Fibrosis Foundation for the use of CF Foundation Patient Registry data.\
- The patients, care providers, and clinic coordinators at CF centers.
- Kristen Miller, MS

>>>>

---
title: "The Changing Landscape of Cystic Fibrosis Related Diabetes Risk Factors Between 2003-2018"
author:
  - name: Tim Vigers
    orcid: 0000-0001-7087-8570
    email: timothy.vigers@cuanschutz.edu
    affiliations:
      - name: University of Colorado Denver, Anschutz Medical Campus
date: "June 17, 2025"
date-format: long
format: 
  revealjs:
    scrollable: true
    embed-resources: true
editor: source
bibliography: wnar_references.bib
csl: american-medical-association.csl
---

@@

- *CFTR* modulator therapies have completely transformed CF care.
- Median predicted survival increased from 34 years in 2004 to 46 years in 2019.[@cysticfibrosisfoundationCysticFibrosisFoundation2024]
- We expect to see even greater gains with the approval of new highly effective *CFTR* modulator therapies. [@Bessonova731; @middletonElexacaftorTezacaftorIvacaftor2019]
- However, despite these advances CFRD remains prevalent and there have been few large-scale studies examining changes in the epidemiology of CFRD over time.[@moranClinicalCareGuidelines2010; @marshallEpidemiologyCysticFibrosisrelated2005; @adlerHyperglycemiaDeathCystic2011; @chamnanDiabetesDeterminantMortality2010]

@@

## The CFF-PR

- These variables were also readily available in the United States Cystic Fibrosis Foundation Patient Registry (CFFPR) and have been collected consistently since 2003.
- The CFF-PR is national database that aims to collect information on everyone with CF who receive care in CF Foundation-accredited care centers.

## The CFF-PR

- In 2023 there were 33,288 people in the registry (about 95%).
- The study P.I. wanted to use annualized data for this analysis.
  - Easier to get permission for this because there are fewer PHI concerns than encounter-level data.
- Very few model selection issues (yay!)

@@

## New Diagnoses in 2018 {.scrolling}

```{r}
#| message: false
load("./Data_Clean/survival_dataset.RData")
annual %>%
  filter(ReviewYear == 2018) %>%
  select(
    CFRD, Age_YrEnd, Gender, Race, Hispanicrace, bmi_perc,
    bmi_cat, MutClass, tube_feeding, liver_disease,
    A_FEV1_pct_predicted, A_FVC_pct_predicted, NumPulmExacerbation,
    A_corticosteroids1, A_IsOnEnzymes, A_Vx770, A_VX809comb
  ) %>%
  tbl_summary(
    by = CFRD, missing_text = "Missing",
    type = list(A_corticosteroids1 ~ "categorical"),
    statistic = list(c(
      "Age_YrEnd", "bmi_perc", "A_FEV1_pct_predicted",
      "A_FVC_pct_predicted"
    ) ~ "{mean} ({sd})")
  ) %>%
  add_overall()
```

@@

Every person in the CFF-PR from 2003-2018

```{r}
#| fig-width: 16
#| fig-height: 9
# KM plot
fit <- survfit(
  Surv(time = tstop, event = as.numeric(CFRD) == 2) ~ age_group,
  data = annual %>%
    group_by(eDWID) %>%
    slice_max(ReviewYear)
)
# Visualize with survminer
ggsurvplot(fit,
  data = annual, risk.table = T, cumevents = T, palette = "simpsons",
  conf.int = T
) + xlab("Days From 2003-01-01")
```

## Data Setup {.smaller}

- Because we don't have an exact date of CFRD diagnosis, the only information we have is that the diagnosis occurred between January 1st of one year and January 1st of the next.
  - Participants had to have two consecutive rows with CFRD, but first instance was considered onset.
- Data are interval censored.
- Intervals are on a discrete grid (i.e. start and stop times are the same for all participants in a given year).
- Data include time-varying covariates such as age group, BMI category, etc.
- We categorized variables such as BMI and FEV1 mostly for the sake of interpretability for a clinical audience.
  - For BMI in particular though, we were not sure that the effect would be linear and categorization allowed for more flexible models.

@@

@@

## Two Good Options

- Joint modeling of longitudinal and survival data.[@hendersonJointModellingLongitudinal2000]
  - The gold standard, do this if you can.
  - Complex, requires somewhat specialized knowledge.
  - Ask Elizabeth for help with this.

- Time-dependent Cox models
  - More appropriate for external covariates (measures that vary independent of the failure time).
  - Or baseline covariates where the effect changes over time.[@zhangTimevaryingCovariatesCoefficients2018]

## Time Dependent Cox Model

@@

## Time Dependent Coefficients

@@

```{r}
#| echo: true
data(cancer)
vet2 <- survSplit(Surv(time, status) ~ .,
  data = cancer, cut = c(90, 180), episode = "tgroup", id = "id"
)
vfit2 <- coxph(Surv(tstart, time, status) ~ age + sex + ph.ecog +
  pat.karno:strata(tgroup), data = vet2)
vfit2
```

@@

```{r}
#| echo: true
annual_split <- survSplit(
  Surv(
    time = tstart, time2 = tstop,
    event = as.numeric(CFRD) == 2
  ) ~ .,
  data = annual, cut = c(2192, 4018), episode = "tgroup"
)
cox_res2 <- coxph(
  Surv(time = tstart, time2 = tstop, event) ~
    age_group:strata(tgroup) + Gender + Race + Hispanicrace + bmi_cat +
    MutClass:strata(tgroup) + A_IsOnEnzymes:strata(tgroup) +
    tube_feeding:strata(tgroup) + liver_disease:strata(tgroup) +
    ppFEV1_cat:strata(tgroup) + NumPulmExacerbation:strata(tgroup) +
    A_corticosteroids1:strata(tgroup),
  data = annual_split, cluster = eDWID
)
```

@@

# An Alternative Approach

@@

- Per the authors of the `survival` package, likelihood equations use only one copy of any subject, and the program picks out the correct row of data at each time.[@therneauUsingTimeDependent2024]
  - This is why we don't need to worry about correlated data even though subjects have multiple observations.
  - Not losing much information by splitting our data up.

@@

@@

- The middle age group is no longer at increased risk compared to the youngest.
- Female and Black people with CF remain at increased risk of developing CFRD.
- People with CF who are in the "undernourished" BMI category are no longer at increased risk, but those who are overweight are.

- Other risk factors appear relatively stable over time.
- Ivcaftor may reduce the risk of CFRD, although combination therapy does not appear to have an effect.
- Time-varying Cox models strike a nice balance between clinical interpretability and statistical rigor.
- Everyone should take a survival class!

@@

- Methods available for interval censored data in the `survival` R package[@therneauPackageSurvivalAnalysis2024] are slightly limited.
  - Other packages such as `icenReg`[@anderson-bergmanIcenRegRegressionModels2017] may be able to fit more flexible models.
- Categorizing variables is not always ideal, and there is some concern about loss of information.
  - Currently we are working on extending this analysis using joint modeling for more flexibility.
    - We will be able to use encounter-level data for this.

- Data end in 2018 but Trikafta (a triple combination therapy of Elexacaftor, Ivacaftor, and Tezacaftor) was approved in 2019.
  - Need to study later generations of CFTR modulators.
- Registry data has natural limitations because each center must track and report their own data.

## Special Thanks

- The Cystic Fibrosis Foundation for the use of CF Foundation Patient Registry data.
- The patients, care providers, and clinic coordinators at CF centers.
- 2025 WNAR/IMS
- Elizabeth Juarez-Colunga, PhD
- Kristen Miller, MS
